---
title: "Biostat 203B Homework 1"
author: "Valentina Arputhasamy"
subtitle: Due Jan 22 @ 11:59PM
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=F}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email.

**Solution: Done**

2. Create a **private** repository `biostat-203b-2021-winter` and add `Hua-Zhou`, `Chris-German` and `ElvisCuiHan` as your collaborators with write permission.

**Solution: Done**

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `master` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `master` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and data sets to reproduce results) in `master` branch.

**Solution: Done**

4. After each homework due date, teaching assistant and instructor will check out your master branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

**Solution: Done**

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

## Q2. Linux Shell Commands

1. This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic-iv.mit.edu/docs/access/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

2. The `/usr/203b-data/mimic-iv/` folder on teaching server contains data sets from MIMIC-IV. Refer to <https://mimic-iv.mit.edu/docs/datasets/> for details of data files.  

    ```{bash}
    ls -l /usr/203b-data/mimic-iv
    ```

Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/usr/203b-data/mimic-iv` directly in following exercises. 

    Use Bash commands to answer following questions.

3. Display the contents in the folders `core`, `hosp`, `icu`. What are the functionalities of the bash commands `zcat`, `zless`, `zmore`, and `zgrep`? 

Contents of the folder `core`:

```{bash}
ls /usr/203b-data/mimic-iv/core
```

Contents of the folder `hosp`:

```{bash}
ls /usr/203b-data/mimic-iv/hosp
```

Contents of the folder `icu`:

```{bash}
ls /usr/203b-data/mimic-iv/icu
```

The bash command `zcat` is used to view the contents of a compressed/zipped file without actually decompressing it.

#The bash command `zless` is used to paginate a compressed file.

##What is the difference between zless and zmore?

#The bash command `zmore` is used to paginate a compressed file.

The bash command `zgrep` is used to search for expressions from a file even if it is compressed (`zgrep` has the same functionalities as `grep`, however unlike `grep` it can be applied to compressed files). 


4. What's the output of following bash script?
    ```{bash, eval=F}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        ls -l $datafile
      done
    ```

The `ls -l` command lists detailed contents of a directory. In the above bash script, the contents of any file ending in `.gz` within the folder `core` are outputted. The output shows the permissions for each file, the owner of the file, the size of the file, the date and time the file was last modified and file/folder name.

Display the number of lines in each data file using a similar loop.

```{bash}
  for datafile in /usr/203b-data/mimic-iv/core/*.gz
    do 
      zcat $datafile | wc -l 
      done
```

5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, `language`, `marital_status`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq`, `wc`, and so on.)

To display the first few lines of `admissions.csv.gz` run the following code:

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | head -10
```

The following code tells us how many rows are in the data file:

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk 'END {print NR}'
```

According to the output, there are 524521 rows in the data file `admissions.csv.gz`.

How many unique patients (identified by `subject_id`) are in this data file?

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | sort $1 | uniq | awk 'END {print NR -1}'
```

According to the output, there are 524520 unique patients in the data file `admissions.csv.gz` (removed the header from the count).

## Admission_type

### Unique Values

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, 'FNR >1 {print $6}' | sort | uniq | head
```
### Count

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, 'FNR >1 {print $6}' | sort | uniq -c | head
```
## Admission Location

### Unique Values

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, 'FNR >1 {print $7}' | sort | uniq | head
```
### Count

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, 'FNR >1 {print $7}' | sort | uniq -c | head
```
## Insurance

### Unique Values

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, 'FNR >1 {print $9}' | sort | uniq | head
```

### Count

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, 'FNR >1 {print $9}' | sort | uniq -c | head
```

## Language

### Unique Values
```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, 'FNR >1 {print $10}' | sort | uniq | head
```

### Count

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, 'FNR >1 {print $10}' | sort | uniq -c | head
```

## Marital Status

### Unique Values

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, 'FNR >1 {print $11}' | sort | uniq -c | head
```

## Ethnicity

### Unique Values

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, 'FNR >1 {print $12}' | sort | uniq | head
```
### Count

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, 'FNR >1 {print $12}' | sort | uniq -c | head
```

## Q3. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    curl http://www.gutenberg.org/cache/epub/42671/pg42671.txt > pride_and_prejudice.txt
    ```

Do **not** put this text file `pride_and_prejudice.txt` in Git. Using a `for` loop, how would you tabulate the number of times each of the four characters is mentioned?

```{bash}
for string in 'Elizabeth' 'Jane' 'Lydia' 'Darcy'
  do grep -o $string pride_and_prejudice.txt | wc -l
done
```

The name `Elizabeth` occurs 634 times, the name `Jane` occurs 294 times, the name `Lydia` occurs 170 times, and the name `Darcy` occurs 417 times. Therefore, the name `Elizabeth` is the name that was most mentioned in the book *Pride and Prejudice*, followed by the name `Darcy`. 

2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```

Upon execution, the first command, 'hello, world' will display in the file test1.txt. If there was any other content in the file test1.txt, it will be replaced with 'hello, world.' 

However, in the second command, 'hello, world' is appended to any existing content in the file test2.txt. If the file test2.txt doesn't exist yet, the command will create a file called test2.txt with 'hello, world' inside.

3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pride_and_prejudice.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

### Q4. More fun with Linux

Try these commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

**Solution** : 

`cal` displays a calendar of the current month.

```{bash}
cal
```

`cal 2021` displays a calendar for each month of the year 2021. 

```{bash}
cal 2021
```

`cal 9 1752` displays a calendar for the 9th month of the year (September) for the year 1972. However, the calendar is missing 11 days (3rd to the 13th). I looked it up, and it's apparently because in September 1972, 11 days were skipped to make up for lack of leap year adjustments.

```{bash}
cal 9 1752
```

`date` displays the day of the week, the date, and the time (Coordinated Universal Time).
```{bash}
date
```

`hostname` is used to view or change a system’s domain and hostname.
```{bash}
hostname
```

