---
title: "Biostat 203B HW3"
author: "Valentina Arputhasamy"
subtitle: Due Mar 5 @ 11:59pm
date: "02/25/2021"
output: html_document
---

```{r}
library(DBI)
library(RPostgreSQL)
library(tidyverse)
library(lubridate)
library(data.table)
```


## Q1. Compile the ICU cohort in HW2 Q8 from the PostgreSQL database `mimiciv`. 

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
is_teaching_server = Sys.info()[["nodename"]] == "ucla-biostat-203b"
```

Below is an outline of steps.

1. Connect to database `mimiciv`. We are going to use username `postgres` with password `postgres` to access the `mimiciv` database.
```{r}
# Load configuration settings
#host  <- '127.0.0.1'
#port  <- '5432'
user  <- ifelse(Sys.info()[["nodename"]] == "ucla-biostat-203b", 
                'postgres', # on teaching server
                'huazhou')  # on my MacBook
password <- ifelse(Sys.info()[["nodename"]] == "ucla-biostat-203b",
                   'postgres', # on teaching server
                   'huazhou')  # on my MacBook   
dbname <- 'mimiciv'
# Connect to the database using the configuration settings
(con <- dbConnect(RPostgreSQL::PostgreSQL(), 
                  dbname = dbname, 
                  user = user, 
                  password = password))
```

2. List all schemas in the `mimiciv` database.
```{r}
dbGetQuery(con, "SELECT SCHEMA_NAME FROM INFORMATION_SCHEMA.SCHEMATA")
```

List tables in the `mimiciv` database:
```{r}
dbListTables(con)
```

List schemas and tables in the `mimiciv` database (bash command).
```{bash, include = !is_teaching_server, eval = F}
psql -U huazhou -d mimiciv -c "\dt *."
```

```{bash, include = is_teaching_server, eval = is_teaching_server}
psql -U postgres -d mimiciv -c "\dt *."
```

3. Connect to the icustays table. Note how to use `Id()` to specify the schema containing the table.
```{r}
icustays_tble <- tbl(con, Id(schema = "mimic_icu", table = "icustays")) %>%
  print(width = Inf)
```

4. Connect to the patients table.
```{r}
# TODO
patients_tble <- tbl(con, Id(schema = "mimic_core", table = "patients")) %>% 
  print(width = Inf)
```

5. Connect to the admissions table.
```{r}
# TODO
admissions_tble <- tbl(con, Id(schema = "mimic_core", table = "admissions")) %>% 
  print(width = Inf)
```

6. Connect to the `mimic_labevents_icu` table.
```{r}
# TODO
labevents_tble <- tbl(con, Id(schema = "public", table = "mimic_labevents_icu")) %>%
  print(width = Inf)
```

7. Connect to `mimic_chartevents_icu` table.
```{r}
# TODO
chartevents_tble <- tbl(con, Id(schema = "public", table = "mimic_chartevents_icu")) %>%
  print(width = Inf)
```

8. Put things together. Using one chain of pipes `%>%` to perform following data wrangling steps: (i) keep only the first ICU stay of each unique patient, (ii) merge in admissions and patients tables, (iii) keep adults only (age at admission >= 18), (iv) merge in the labevents and chartevents tables, (v) display the SQL query, (vi) collect SQL query result into memory as a tibble, (vii) create an indicator for 30-day mortality, (viii) save the final tibble to an `icu_cohort.rds` R data file in the `mimiciv_shiny` folder.

```{r}
# make a directory mimiciv_shiny
if (!dir.exists("mimiciv_shiny")) {
  dir.create("mimiciv_shiny")
}
```

```{r}
# TODO
icustays_tble$intime <- ymd_hms(icustays_tble$intime)

icustays_tble %>% 
  group_by(subject_id) %>%
  filter(intime == min(intime)) %>% 
  ungroup() %>%
  left_join(admissions_tble, by = c("subject_id", "hadm_id")) %>%
  left_join(patients_tble, by = "subject_id") %>%
  mutate(age_at_adm = year(admittime) - anchor_year + anchor_age) %>%
  filter(age_at_adm >= 18) %>%
  left_join(labevents_tble, by = c("subject_id", "hadm_id")) %>% 
  left_join(chartevents_tble, by = c("subject_id", "hadm_id", "stay_id")) %>%
  show_query() %>%
  as_tibble() %>%
  collect() %>%
  mutate(days.til.death = ceiling((((deathtime - admittime)/60)/60)/24),
         died.within.30 = ifelse(days.til.death <= 30, 1, 0)) %>%
  saveRDS(file = "mimiciv_shiny/icu_cohort.rds")
```

Close database connection and clear workspace.
```{r}
dbDisconnect(con)
rm(list = ls())
```

## Q2. Shiny app

Develop a Shiny app for exploring the ICU cohort data created in Q1. The app should reside in the `mimiciv_shiny` folder. The app should provide easy access to the graphical and numerical summaries of variables (demographics, lab measurements, vitals) in the ICU cohort.
